$p$-adic precision

### 2014-02-15 padicprec.tex

\newcommand{\lb}{\ensuremath{\llbracket}}
\newcommand{\rb}{\ensuremath{\rrbracket}}

@@

Let $K$ be a non-Archimedean local field with absolute value
$|\cdot| : K \to \R_+$ and value group $p^\Z$.   For example, we may take $K = \Qp$ with the
$p$-adic absolute value or $K = \Fp\lb t \rb$ with the $t$-adic absolute value.
Write $\OK$ for the ring of integers of $K$ and choose a dense subring $R \subset K$
consisting of elements that can be represented exactly.  For $K = \Qp$ we may
choose $R = \Z[\frac1p]$ or $R = \Q$; for $K = \Fp\lb t \rb$ we may choose
$R = \Fp[t, t^{-1}]$ or $R = \Fp(t)$.  Choose a collection of subsets $S_h \subset R$
indexed by positive integers $h$ so that
\begin{enumerate}
\item $S_h \subset S_{h+1}$,
\item $\{B_K(x, p^{-h}) : x \in S_h\}$ disjointly covers $K$.
\end{enumerate}
The choice of the $S_h$ is possible since $K$ is non-Archimedean and locally compact.

An \emph{absolute value} on a field $K$ is a multiplicative map $|\cdot| 
: K \to \R_+$ satisfying the triangle inequality and sending every 
nonzero element of $K$ to a strictly positive real number. It is said 
\emph{ultrametric} if it satisfies the following strong from of the 
triangle inequality:
$$|x+y| \leq \max(|x|, |y|).$$
Given an absolute value $|\cdot|$ on $K$, we note that $|K^\times|$ is a 
multiplicative subgroup of $\R_+^\times$. It is then either discrete or 
dense. We shall say that $|\cdot|$ is \emph{discrete} if $|K^\times|$ is
discrete. The valued field $(K, |\cdot|)$ is said \emph{complete} if all 
Cauchy sequences (with respect to the distance $d$ defined by $d(x,y) = 
|x-y|$) converge.

The formula $|x| = p^{-\val(x)}$ defines an ultrametric absolute value on 
the field $\Qp$ for which it is complete. More generally, the $p$-adic 
absolute value extends uniquely to any finite extension $K$ of $\Qp$ and 
turns $K$ into a complete ultrametric field. Another classical example of 
complete ultrametric field is the field of Laurent series over a finite
field, the absolute value being defined from the valuation of a series.

From now, we fix a field $K$ equipped with an ultrametric absolute value 
denoted by $|\cdot|$ and we assume that $K$ is complete. Given $v$ a 
$K$-vector space (of possibly infinite dimension), an \emph{ultrametric 
norm} on $V$ is a map $\Vert\cdot\Vert : V \to \R^+$ satisfying:
\begin{enumerate}[(i)]
\item $\Vert x\Vert = 0$ if and only if $x = 0$;
\item $\Vert \lambda x\Vert = |\lambda| \cdot \Vert x\Vert$;
\item $\Vert x+y\Vert \leq \max(\Vert x\Vert, \Vert y\Vert)$.
\end{enumerate}
A norm gives rise to a distance and a normed $K$-vector space $V$ is 
said \emph{complete} when any Cauchy sequence with values in $V$ 
converges. A normed $K$-vector space which is complete is called a 
\emph{$K$-Banach}.

Similarly to the real case, any finite dimensional normed $K$-vector 
space is complete and all norms over such a space are equivalent 
(\emph{i.e.} they define the same topology). Many usual Theorems 
concerning Banach spaces (\emph{e.g.} open mapping Theorem, closed graph 
Theorem) remain true in the ultrametric setting.

\begin{deftn} \label{deftn : diff}
Let $V, W$ be two normed $K$-vector spaces, let $U \subset V$ be an open 
subset of $V$ and let $f : U \rightarrow W$ be a map. Then $f$ is called 
\emph{differentiable} at $v_0 \in U$ if there exists a 
continuous linear map $f'(v_0) : U \rightarrow W$ such that for any 
$\varepsilon >0$, there exists an open neighborhood $U_\varepsilon 
\subset U$ containing $v_0$ with:
\[ 
\forall v, w \in U_\varepsilon, \quad
\Vert f(v)-f(w)-f'(v_0) \cdot \left( v-w \right) \Vert 
\leq \varepsilon \Vert v-w \Vert. 
\]
The linear map $f'(v_0)$ is called the \emph{differential} of $f$ at $v_0$.
\end{deftn}

One can derive all usual properties of the differential from this 
definition: unicity, chain rules, \emph{etc}.

Let $E$ be a $K$-normed vector space. Given $r \in \R^+$, 
we denote by $B^-_E(r)$ (resp. by $B_E(r)$) the set of elements $x 
\in E$ such that $\Vert x \Vert < r$ (resp. $\Vert x \Vert \leq r$). 
We insist on the following maybe unusual fact: $B^-_E(r)$ and 
$B_E(r)$ are both open and closed for the norm topology on $E$.
Clearly, $B_K(0,1)$ is stable under multiplication. Moreover, the 
ultrametric triangle inequality ensures that $B_K(1)$ is stable under 
addition as well. Thus, $B_K(1)$ is a subring of $K$: it is often 
called the \emph{integer ring} of $K$.

\begin{deftn}
Let $E$ be a $K$-Banach.
A \emph{lattice} in $E$ is an open bounded sub-$B_K(1)$-module of $E$.
\end{deftn}

\begin{rmk}
Any lattice $H$ in $E$ is also closed since its complement is the
union of all cosets $a + H$ (with $a \not\in H$) which are all open.
\end{rmk}

@@

>>>>

\newcommand{\lb}{\ensuremath{\llbracket}}
\newcommand{\rb}{\ensuremath{\rrbracket}}
\newcommand{\lp}{(\!(}
\newcommand{\rp}{)\!)}

@@

Let $K$ be a field with absolute value $|\cdot| : K \to \R_{\ge 0}$. 
We assume that the induced metric is an ultrametric -- 
\[ \lvert x + y \rvert \leq \max(\lvert x \rvert, \lvert y \rvert) \]
-- and that $K$ is complete with respect to it.
For example, we may take $K = \Qp$ with the
$p$-adic absolute value or $K = k\lp t \rp$ with the $t$-adic absolute value.
Write $\OK$ for the ring $\{x \in K : \lvert x \rvert \le 1\}$ and assume
that $K$ contains a dense subring $R \subset K$
consisting of elements that can be represented with a finite amount of space.
For $K = \Qp$ we may choose $R = \Z[\frac1p]$ or $R = \Q$;
for $K = \Fp\lb t \rb$ we may choose $R = \Fp[t, t^{-1}]$ or $R = \Fp(t)$.

If $E$ is a $K$-vector space, possibly of infinite dimension, then an
\emph{ultrametric norm} on $E$ is a map $\Vert\cdot\Vert : E \to \R^+$ satisfying:
\begin{enumerate}[(i)]
\item $\Vert x\Vert = 0$ if and only if $x = 0$;
\item $\Vert \lambda x\Vert = |\lambda| \cdot \Vert x\Vert$;
\item $\Vert x+y\Vert \leq \max(\Vert x\Vert, \Vert y\Vert)$.
\end{enumerate}
A \emph{$K$-Banach space} is a complete normed $K$-vector space.

\begin{deftn} % Separated definitions?
Let $E$ be a $K$-Banach space.
A \emph{lattice} in $E$ is an open bounded sub-$\OK$-module of $E$.
\end{deftn}

For a $K$-Banach space $E$ and $r \in \R_{\ge 0}$, write
\begin{align*}
B_E(r) &= \{ x \in E : \Vert x \Vert \le r \}, \\
B^-_E(r) &= \{ x \in E : \Vert x \Vert < r\}.
\end{align*}
Note that $B_E(r)$ and $B^-_E(r)$ are both lattices.

\begin{deftn}
Suppose $E$ is a $K$-Banach space and $I$ a set.
A family $(x_i)_{i \in I} \subset E$ is a \emph{Banach basis} for $E$ if
every element $x \in E$ can be written
\[
x = \sum_{i \in I} \alpha_i x_i
\]
for scalars $\alpha_i \in K$ with $\alpha_i \to 0$, and $\Vert x \Vert = \sup_{i \in I} \lvert \alpha_i \rvert$.
\end{deftn}
Note that if $E$ is finite dimensional then the condition $\alpha_i \to 0$ is vacuous.
% Lattice from Banach basis

\subsection{Approximate elements}

Suppose that $E$ is a $K$-Banach space with basis $(x_i)_{i \in I}$.  We say that an element
$x \in E$ is \emph{exact} if there is a finite subset $J \subseteq I$ and scalars $\alpha_j \in R$ with
\begin{equation} \label{eq:exact_elt}
x = \sum_{j \in J} \alpha_j x_j.
\end{equation}
The set of exact elements is dense in $E$.

An approximate element is a pair $(x, H)$ where $x \in E$ is an exact element
and $H$ is a lattice in $E$.  Such a pair represents an undetermined element of the coset
$x + H$.  We will frequently write $x + O(H)$ to emphasize the fact that $H$ represents
the uncertainty in the value of the approximate element.  In the special case that $E = K = \Qp$,
we recover the standard notation $a + O(p^n)$ for an approximate $p$-adic element.

Note that using an arbitrary lattice to represent the precision of an approximate element
can reduce precision loss when compared to storing the precision of each coefficient
$\alpha_i$ in \eqref{eq:exact_elt} separately.  Consider the function 
$f : \Qp^2 \to \Qp^2$ mapping $(x,y)$ to $(x+y, x-y)$ and the problem of computing
$f \circ f(a + O(p^n), b + O(p^m))$.

@@

%\subsection{Ultrametric analysis}
%\label{ssec:ultrametric}

%Similarly to the real case, any finite dimensional normed $K$-vector 
%space is complete and all norms over such a space are equivalent 
%(\emph{i.e.} they define the same topology). Many usual Theorems 
%concerning Banach spaces (\emph{e.g.} open mapping Theorem, closed graph 
%Theorem) remain true in the ultrametric setting.

%\begin{deftn} \label{deftn : diff}
%Let $V, W$ be two normed $K$-vector spaces, let $U \subset V$ be an open 
%subset of $V$ and let $f : U \rightarrow W$ be a map. Then $f$ is called 
%\emph{differentiable} at $v_0 \in U$ if there exists a 
%continuous linear map $f'(v_0) : U \rightarrow W$ such that for any 
%$\varepsilon >0$, there exists an open neighborhood $U_\varepsilon 
%\subset U$ containing $v_0$ with:
%\[ 
%\forall v, w \in U_\varepsilon, \quad
%\Vert f(v)-f(w)-f'(v_0) \cdot \left( v-w \right) \Vert 
%\leq \varepsilon \Vert v-w \Vert. 
%\]
%The linear map $f'(v_0)$ is called the \emph{differential} of $f$ at $v_0$.
%\end{deftn}

%One can derive all usual properties of the differential from this 
%definition: unicity, chain rules, \emph{etc}.

%The following Lemma shows that the notion of lattices behaves rather 
%well under linear\footnote{We shall see in the sequel (\emph{cf} Lemma 
%\ref{lem:main}) that, under some extra hypothesis, the linearity 
%assumption can be drastically relaxed.} transformations.

%\begin{lem}
%\label{lem:morlat}
%Let $E$ and $F$ be two $K$-Banach and $f : 
%E \to F$ be a continuous \emph{surjective} linear map.
%If $H$ is a lattice in $E$, then $f(H)$ is a lattice in $F$.
%\end{lem}

%\begin{proof}
%It is clear that $f(H)$ is a module over $B_K(1)$ which is bounded.
%The fact that it is open is a direct consequence of the open mapping
%Theorem.
%\end{proof}

%To each lattice $H$ in $E$, one can attach a norm $\Vert \cdot \Vert_H$ 
%on $E$ defined as follows. We pick $x \in E$, $x \neq 0$. We introduce
%the set $I_x$ consisting of scalars $\lambda \in K$ such that $\lambda x 
%\in H$. It follows from the fact that $H$ is open and bounded in $E$
%that $I_x$ is nonempty and bounded in $K$. Therefore, the image $|I_x|$ 
%of $I_x$ under the absolute value map is bounded and contains a positive
%element. We can then define:
%$$\Vert x \Vert_H = (\sup |I_x|)^{-1}.$$
%It is an exercise to check that $\Vert \cdot \Vert_H$ is a norm over 
%$E$.

%\begin{lem} \label{lem:reductionball}
%\begin{enumerate}
%\item The norms $\Vert \cdot \Vert_H$ and $\Vert \cdot \Vert$ are 
%equivalent.
%\item Let $B^-_{E,H}(1)$ (resp. by $B_{E,H}(1)$) the set of elements $x 
%\in E$ such that $\Vert x \Vert_H < 1$ (resp. $\Vert x \Vert_H \leq 1$).
%Then:
%$$B^-_{E,H}(1) \subset H \subset B_{E,H}(1)$$
%and the last inclusion is an equality if the absolute value on $K$ is
%discrete.
%\end{enumerate}
%\end{lem}

%\begin{proof}
%Left to the reader.
%\end{proof}
